---
title: "Spatio-Temporal Analysis of America temperature in 2015"
author: "Zhu Lin"
output: pdf_document
---


# Getting startted
The film "the wandering earth", which has attracted a lot of attention recently, sets up an extremely harsh climate which is close to collapse after a great change of climate. Such a catactrophic scenario is not just a figment of the bad weather that is common in many places. Climate change and potential global warming have become the significant issue of the day, monitoring temperaturesacross the region, therefore, is becoming increasingly important in response to weather-related disasters.

The data files US_weather_2015 and weather_stations correspondingly contain temperature of the United States at different stations and and their locations (marked by longitude, latitude and elevation) from January 1 to December 31, 2015, a period of 365 days. Given spatial statistics models generally only take longitude and latitude into consideration, we substract the variable elevation (elev) and replace the temperature with the mean temperature(temp_mean) in the same time at the same location (including stn, lon and lat). Furthermore, to facilitate analysis, we should merge two datasets according to the same variable station number (stn). Top of all, we assume that temporal resolution is one day, and hence we are expected to divide the whole data into 365 sub dataframes including the information of each day, and denote them by "day1", "day2" and so on. 

# Exploratory data analysis
## Data description
The following packages are used in this project. 

```{r, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(maps)
```

Read two data files US_weather_2015 and weather_stations

```{r, warning=FALSE}
weather = read.csv("US_weather_2015.csv")
loc = read.csv("weather_stations.csv")
```

Dataframe weather has five variables: stn, year, month, day and temp.
```{r, warning=FALSE}
head(weather)
```

Dataframe loc has seven variables: stn, name, country, state, lat, lon and elev.
```{r, warning=FALSE}
head(loc)
```

Merge weather and loc by common variables stn, lat and lon and remove duplicate rows based on all variables
```{R, eval = F}
## merge weather and location(loc)
library(dplyr)
loc = select(loc,-c(name,country,state,elev))
data = left_join(weather, loc, by = "stn")
data = filter(data, is.na(lat) == F)
## Remove Duplicate Rows based on all variables
data = distinct(data)
```

Add some variables, do data cleaning and replace "temperature" with "temperature_mean".
```{R, eval = F}
## add a variable "day" to identify date
data = mutate(data, num = as.numeric(ISOdate(2005,month,day)-ISOdate(2005,1,1),
                                   units="days")+1)
## add a variable "atr" to distinguish their attributes
data = mutate(data, attr = (stn/10000+lon/10+lat+num)*10000)
temp_mean = summarise(group_by(data, attr), temp_mean = mean(temp))
data_clean = left_join(data, temp_mean, by = "attr")
data_clean = select(data_clean,-c(temp))
data_clean = distinct(data_clean)
```

To familiarize ourselves with the geography of the dataset, we will initially ignore the temporal component of the dataset and examine the spatial distribution of temperatures on a single day. Take day 1 as an example, therefore, we extract data of January 1, 2015 and add variables lon_rank and lat_rank.
```{R, eval = F}
## Extract day 1's data and sort it
day1 = filter(data_clean, num == 1)
day1 = arrange(day1, lon, lat)
day1 = mutate(day1, lon_rank = min_rank(lon), lat_rank = min_rank(lat))
```

Since the temperature points are discrete with respect to the whole space (in the longitude-latitude-axis matrix, only 0.02% of elements are not NA), it is not available to show the United States temperatures by using the image.plot command in the fields package. In this case, we introduce the following method to plot the discrete points' temperature distribution image.

```{R, eval = F}
library(ggplot2)
library(maps)

statesMap <- map_data("state")

# Extract the states in the main US territory
states = unique(statesMap$region)
# match locations in day1 with regions in statesMap 
tempdata=filter(day1, lat < 50 & lat > 25 & lon > (-125) & lon < (-65))

ggplot(data = statesMap) + 
  geom_polygon(aes(x = long, y = lat, group = group), color = "black", fill="white") + 
  coord_fixed(1.3) +
  guides(fill=FALSE)+  # do this to leave off the color legend
  geom_point(mapping=aes(lon,lat, colour = temp_mean), data = tempdata, size = 0.7) +
  scale_colour_gradient2(low = "blue", high = "red", mid="lightblue", midpoint = 32)
```

![Temperature data of the United States observed on January 1 in 2015]("UStemp")

A pronounced temperature gradient is visible from highs of over 85.6 degrees Celsius in the north of the study area which is near to the equator to a low of -14.7 degrees Celsius towards the southern boundary. This is not only indicative of spatial correlation in the dataset, but it also shows that the data are not stationary, as the mean temperature must vary strongly with latitude. A more precise conclusion can be reached by combining Figure 1 with Figure 2.

![ Histogram of US temperature on January 1 in 2015]("dist1")

According to Figure 1 and Figure 2, we can find that on January 1 in 2015, the temperature in most parts of the United States was concentrated at 30 degrees Celsius and extreme temperatures only account for a small proportion.

![ Mean of US temperature in latitude  on January 1 in 2015]("m_lat")

![ Variance of US temperature in latitude  on January 1 in 2015]("var_lat")

![ Mean of US temperature in longitude  on January 1 in 2015]("m_lon")

![ Variance of US temperature in longitude  on January 1 in 2015]("var_lon")

![ Mean of US temperature by date]("m_date")

![ Variance of US temperature by date]("var_date")

# Research objectives
In light to Figure 3 - Figure 8, we can confirm that there are some trends in latitude, longitude and date marginally. For simplicity, in variogram modeling and kriging throughout this paper, we will treat the latitude and longitude coordinates as if they are Cartesian. In geostatistics, we generally consider the temperature is normally distributed with mean and covariance structure being functions of latitude, longitude and time. Furthermore, the mean function may be the linear combination or other nonlinear forms of three factors. Apart from this, the covariance may introduce the Matern class, which is popular in spatial statistics. Considering these cases, we will figure out the structure of the normal distribution by the following steps.

1. Construct spatial data analysis regardless of the influence of time and find the most appropriate model.
2. Construct Spatial-Temporal analysis.
3. Various forms of Kriging can be used to attempt to fill gaps caused by orbital clipping and cloud cover so that we can obtain the complete temperature variation diagram across the United States. 
